---
title: "R Plots"
author: "Data Knights"
date: "November 7, 2018"
output:
  html_document:
    theme: united
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: no
  github_document:
    toc: yes
---
* * *

## R Plots with Description
    ```{r setup, echo=FALSE, message=FALSE, warning=FALSE, results='hide'} 
    knitr::opts_chunk$set(cache=TRUE)
    knitr::opts_knit$set(root.dir = "C:/Users/ramas/Documents/ISQA 8086/R Plots") 
    ```
#### Research Questions
* How is Bird Banding Data Changing over time? (Changes in the number of sightings)
* Are the breeding and molting seasons changing?
* Has the sex Ratio for birds changed significantly, and does it have a trend?
* Has the number of birds seen each year been stable or trending in a certain direction?
* Has been any change in the wing to weight ratio of the birds over past 20 years?
* Finding the seasonal patterns in the weight changes in the birds.
* Finding the seasonal patterns in Wing to Weight Ratio changes in birds.
    
#### Reading the Data File
* Read the data file. The file cleanData.csv is read and stored in a Dataframe called BandingData.
    ```{r}
    BandingData <- read.csv("cleanData.csv", header = TRUE, stringsAsFactors = FALSE)

    ```

    ```{r, echo=FALSE, warning=FALSE}
    #Recoding year
    BandingData$year[BandingData$year == 0] <- 2000
    BandingData$year[BandingData$year == 1] <- 2001
    BandingData$year[BandingData$year == 3] <- 2003
    BandingData$year[BandingData$year == 4] <- 2004
    BandingData$year[BandingData$year == 5] <- 2005
    BandingData$year[BandingData$year == 6] <- 2006
    BandingData$year[BandingData$year == 7] <- 2007
    BandingData$year[BandingData$year == 8] <- 2008
    BandingData$year[BandingData$year == 9] <- 2009
    BandingData$year[BandingData$year == 10] <- 2010
    BandingData$year[BandingData$year == 11] <- 2011
    BandingData$year[BandingData$year == 12] <- 2012
    BandingData$year[BandingData$year == 13] <- 2013
    BandingData$year[BandingData$year == 14] <- 2014
    BandingData$year[BandingData$year == 15] <- 2015
    BandingData$year[BandingData$year == 16] <- 2016
    BandingData$year[BandingData$year == 96] <- 1996
    BandingData$year[BandingData$year == 97] <- 1997
    BandingData$year[BandingData$year == 98] <- 1998
    
    # Creating subset for Common Yellowthroat species
    coye <- subset(BandingData, SpeciesCode=="COYE")
    
    # Creating subset for Song Sparrow species
    sosp <- subset(BandingData, SpeciesCode=="SOSP")
    
    # Creating subset for Bewick's Wren species
    bewr <- subset(BandingData, SpeciesCode=="BEWR")
    
    # Creating subset for  Bushtit species
    bush <- subset(BandingData, SpeciesCode=="BUSH")
    
    # Creating subset for Chestnut-backed Chickadee species 
    cbch <- subset(BandingData, SpeciesCode=="CBCH")
    
    
    
    library(plyr)
    
    # Number of Male and Female birds spotted every year
    sosp_count_year=count(sosp,c('year','sexRatio'))
    coye_count_year=count(coye,c('year','sexRatio'))
    bewr_count_year=count(bewr,c('year','sexRatio'))
    bush_count_year=count(bush,c('year','sexRatio'))
    cbch_count_year=count(cbch,c('year','sexRatio'))
    
    # comparision of gender ratio
    library(tidyr)
    library(ggplot2)
    ```
    
#### Change in the Sex Ratio of Different species over the years
* **Song Sparrow**: We cannot quite comment on the Sex Ratio of Song Sparrows as there are a lot of NA values, which means the sex of the bird was not determined at the time of capture. However, we can see that population of male birds is higher that female birds for majority of the years.  

    ```{r, echo=FALSE}
    
    # Change in the Sex Ratio of Different species over the years

    sosp_count_year$sexRatio[sosp_count_year$sexRatio == 0] <- "F"
    sosp_count_year$sexRatio[sosp_count_year$sexRatio == 1] <- "M"
    
    
    ggplot(data=sosp_count_year, aes(x=year, y=freq, fill=sexRatio)) +
      geom_bar(stat="identity", position=position_dodge())+xlab("Year")+ylab("Count")+
      labs(title="Change in Sex Ratio of Song Sparrow",fill="Sex")+
      theme(plot.title = element_text(hjust = 0.5)
            )
    ########################################################################################
    ```

* **Common Yellowthroat**: The number of Male birds is significantly higher than that of Female birds and is being constant over the years.

    ```{r, echo=FALSE}

    coye_count_year$sexRatio[coye_count_year$sexRatio == 0] <- "F"
    coye_count_year$sexRatio[coye_count_year$sexRatio == 1] <- "M"
    
    ggplot(data=coye_count_year, aes(x=year, y=freq, fill=sexRatio)) +
      geom_bar(stat="identity", position=position_dodge())+xlab("Year")+ylab("Count")+
      labs(title="Change in Sex Ratio of Common Yellowthroat",fill="Sex")+
      theme(plot.title = element_text(hjust = 0.5))
    
    ######################################################################################
    ```
    
* **Bewick's Wren**: Cannot identify trends as more than 90% of the data is unknown.

    ```{r, echo=FALSE}
    
    bewr_count_year$sexRatio[bewr_count_year$sexRatio == 0] <- "F"
    bewr_count_year$sexRatio[bewr_count_year$sexRatio == 1] <- "M"
    
    ggplot(data=bewr_count_year, aes(x=year, y=freq, fill=sexRatio)) +
      geom_bar(stat="identity", position=position_dodge())+xlab("Year")+ylab("Count")+
      labs(title="Change in Sex Ratio of Bewick's Wren",fill="Sex")+
      theme(plot.title = element_text(hjust = 0.5))
    
    ######################################################################################
    ```
    
* **Bushtit**: The number of Female birds is higher compare to the number of Male birds and has not changed in 20 years.

    ```{r, echo=FALSE}
    
    bush_count_year$sexRatio[bush_count_year$sexRatio == 0] <- "F"
    bush_count_year$sexRatio[bush_count_year$sexRatio == 1] <- "M"
    
    ggplot(data=bush_count_year, aes(x=year, y=freq, fill=sexRatio)) +
      geom_bar(stat="identity", position=position_dodge())+xlab("Year")+ylab("Count")+
      labs(title="Change in Sex Ratio of Bushtit",fill="Sex")+
      theme(plot.title = element_text(hjust = 0.5))
    
    ######################################################################################
    ```
    
* **Chestnut-backed Chickadee**: Cannot identify trends as more than 90% of the data is unknown.

    ```{r, echo=FALSE}
    
    cbch_count_year$sexRatio[cbch_count_year$sexRatio == 0] <- "F"
    cbch_count_year$sexRatio[cbch_count_year$sexRatio == 1] <- "M"
    
    ggplot(data=cbch_count_year, aes(x=year, y=freq, fill=sexRatio)) +
      geom_bar(stat="identity", position=position_dodge())+xlab("Year")+ylab("Count")+
      labs(title="Change in Sex Ratio of Chestnut-backed Chickadee",fill="Sex")+
      theme(plot.title = element_text(hjust = 0.5))
    
    ######################################################################################
    
    ```
    
#### Number of Sightings of different species over the years
* It can be noticed that the number of sightings of Song sparrows is significantly higher than that of other species but has dropped rapidly between 2010 and 2016 which might be a reason to be concerned. The sightings of Bewick's Wren is increasing slowly while the other threee species are fairly constant.

    ```{r, echo=FALSE}
    count_sosp_year = count(sosp, c('year'))
    count_coye_year = count(coye,  c('year'))
    count_bewr_year = count(bewr, c('year'))
    count_bush_year = count(bush,  c('year'))
    count_cbch_year = count(cbch,  c('year'))
    
    ggplot() + 
      geom_line(data=count_sosp_year, aes(x=year, y=freq, color="Song sparrow"),size=1) +
      geom_line(data=count_coye_year, aes(x=year, y=freq, color="Common Yellow Throat"),size=1)+
      geom_line(data=count_bewr_year, aes(x=year, y=freq, color="Bewick's Wren"),size=1) +
      geom_line(data=count_bush_year, aes(x=year, y=freq, color="Bushtit"),size=1)+
      geom_line(data=count_cbch_year, aes(x=year, y=freq, color="Chestnut-backed Chickadee"),size=1)+xlab("Year")+ylab("Count")+labs(title="Sightings by Year",colour="Species Name")+
      theme(plot.title = element_text(hjust = 0.5))
    ```
    
#### Number of Sightings of different Species by month over the span of 5 years
* After analyzing these plots, we found a trend that number of sightings is significantly higher in the summer than compared to the cold seasons for all the species.

    ```{r, echo=FALSE}
    count_sosp = count(sosp, c('month','year'))
    count_coye = count(coye,c('month','year'))
    count_bewr = count(bewr, c('month','year'))
    count_bush = count(bush, c('month','year'))
    count_cbch = count(cbch, c('month','year'))
    
    a <- subset(count_sosp, year=="2016"|year=="2015"|year=="2014"|year=="2013"|year=="2012")
    a$year<-factor(a$year)
    ggplot(data = a, aes(x = month, y = freq, color=year)) + 
      geom_line()+ geom_point()+
      xlab("Month")+ylab("Count")+labs(title="Sightings per month for Song Sparrow",colour="Year")+
      theme(plot.title = element_text(hjust = 0.5))+scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12))
    
    b <- subset(count_coye, year=="2016"|year=="2015"|year=="2014"|year=="2013"|year=="2012")
    b$year<-factor(b$year)
    ggplot(data = b, aes(x = month, y = freq, color=year)) + 
      geom_line()+ geom_point()+
      xlab("Month")+ylab("Count")+labs(title="Sightings per month for Common Yellowthroat",colour="Year")+
      theme(plot.title = element_text(hjust = 0.5))+scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12))
    
    c <- subset(count_bewr, year=="2016"|year=="2015"|year=="2014"|year=="2013"|year=="2012")
    c$year<-factor(c$year)
    ggplot(data = c, aes(x = month, y = freq, color=year)) + 
      geom_line()+ geom_point()+
      xlab("Month")+ylab("Count")+labs(title="Sightings per month for Bewick's Wren",colour="Year")+
      theme(plot.title = element_text(hjust = 0.5))+scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12))
    
    d <- subset(count_bush, year=="2016"|year=="2015"|year=="2014"|year=="2013"|year=="2012")
    d$year<-factor(d$year)
    ggplot(data = d, aes(x = month, y = freq, color=year)) + 
      geom_line()+ geom_point()+
      xlab("Month")+ylab("Count")+labs(title="Sightings per month for Bushtit",colour="Year")+
      theme(plot.title = element_text(hjust = 0.5))+scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12))
    
    e <- subset(count_cbch, year=="2016"|year=="2015"|year=="2014"|year=="2013"|year=="2012")
    e$year<-factor(e$year)
    ggplot(data = e, aes(x = month, y = freq, color=year)) + 
      geom_line()+ geom_point()+
      xlab("Month")+ylab("Count")+labs(title="Sightings per month for Chestnut-backed Chickadee",colour="Year")+
      theme(plot.title = element_text(hjust = 0.5))+scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12))
    ```
    
#### Change in the average weight of each species over the span of 20 years
* The average weight of all the species is constant over the past 20 years.

    ```{r, echo=FALSE, message=FALSE, warning=FALSE}
    #summary statistics based on years for Common Yellowthroat
    detach(package:plyr)
    library(dplyr)
    
    f =coye%>%
      group_by(year)%>% 
      summarise(Mean=mean(cleanWeight, na.rm=TRUE),Max=max(cleanWeight, na.rm=TRUE), Min=min(cleanWeight, na.rm=TRUE), Median=median(cleanWeight, na.rm=TRUE), Std=sd(cleanWeight, na.rm=TRUE))
    
    #summary statistics based on years for Song Sparrow
    g=sosp%>%
      group_by(year)%>% 
      summarise(Mean=mean(cleanWeight, na.rm=TRUE),Max=max(cleanWeight, na.rm=TRUE), Min=min(cleanWeight, na.rm=TRUE), Median=median(cleanWeight, na.rm=TRUE), Std=sd(cleanWeight, na.rm=TRUE))
    
    #summary statistics based on years for Bewick's Wren
    h=bewr%>%
      group_by(year)%>% 
      summarise(Mean=mean(cleanWeight, na.rm=TRUE),Max=max(cleanWeight, na.rm=TRUE), Min=min(cleanWeight, na.rm=TRUE), Median=median(cleanWeight, na.rm=TRUE), Std=sd(cleanWeight, na.rm=TRUE))
    
    #summary statistics based on years for Bushtit
    i=bush%>%
      group_by(year)%>% 
      summarise(Mean=mean(cleanWeight, na.rm=TRUE),Max=max(cleanWeight, na.rm=TRUE), Min=min(cleanWeight, na.rm=TRUE), Median=median(cleanWeight, na.rm=TRUE), Std=sd(cleanWeight, na.rm=TRUE))
    
    #summary statistics based on years for Chestnut-backed Chickadee
    j=cbch%>%
      group_by(year)%>% 
      summarise(Mean=mean(cleanWeight, na.rm=TRUE),Max=max(cleanWeight, na.rm=TRUE), Min=min(cleanWeight, na.rm=TRUE), Median=median(cleanWeight, na.rm=TRUE), Std=sd(cleanWeight, na.rm=TRUE))
    
    ##################################################################################################################
    
    ggplot() + 
      geom_line(data=f, aes(x=year, y=Mean, color="Song sparrow"),size=1) +
      geom_line(data=g, aes(x=year, y=Mean, color="Common Yellow Throat"),size=1)+
      geom_line(data=h, aes(x=year, y=Mean, color="Bewick's Wren"),size=1) +
      geom_line(data=i, aes(x=year, y=Mean, color="Bushtit"),size=1)+
      geom_line(data=j, aes(x=year, y=Mean, color="Chestnut-backed Chickadee"),size=1)+xlab("Year")+ylab("Mean of Weights")+labs(title="Weight changes by Year",colour="Species Name")+
      theme(plot.title = element_text(hjust = 0.5))
    ```
    
#### Change in the average weight of each species by month over the span of 5 years
* The pattern of changes in the weight per month is different for each species, but is fairly constant with a few outliers between 2012 to 2016 

    ```{r, echo=FALSE}
    #mean of weights by month for each year for Common Yellowthroat
    coye_mean = data.frame()
    
    for ( yr in unique(coye$year) )
    {
      for ( mn in unique(coye$month) )
      {
        tmp = subset(coye, year==yr & month==mn)
        mean = mean(tmp$cleanWeight, na.rm=TRUE)
        coye_mean = rbind(coye_mean, data.frame(yr,mn, mean))
      }
    }
    
    coye_mean <- coye_mean[with(coye_mean, order(mn,yr)), ]
    
    ###################################################################################################
    
    #mean of weights by month for each year for Song Sparrow
    sosp_mean = data.frame()
    
    for ( yr in unique(coye$year) )
    {
      for ( mn in unique(coye$month) )
      {
        tmp1 = subset(sosp, year==yr & month==mn)
        mean1 = mean(tmp1$cleanWeight, na.rm=TRUE)
        sosp_mean = rbind(sosp_mean, data.frame(yr,mn, mean1))
      }
    }
    
    sosp_mean <- sosp_mean[with(sosp_mean, order(mn,yr)), ]
    
    ###################################################################################################
    
    #mean of weights by month for each year for Bewick's Wren
    bewr_mean = data.frame()
    
    for ( yr in unique(bewr$year) )
    {
      for ( mn in unique(bewr$month) )
      {
        tmp2 = subset(bewr, year==yr & month==mn)
        mean2 = mean(tmp2$cleanWeight, na.rm=TRUE)
        bewr_mean = rbind(bewr_mean, data.frame(yr,mn, mean2))
      }
    }
    
    bewr_mean <- bewr_mean[with(bewr_mean, order(mn,yr)), ]
    
    ###################################################################################################
    
    #mean of weights by month for each year for Bushtit
    bush_mean = data.frame()
    
    for ( yr in unique(bush$year) )
    {
      for ( mn in unique(bush$month) )
      {
        tmp3 = subset(bush, year==yr & month==mn)
        mean3 = mean(tmp3$cleanWeight, na.rm=TRUE)
        bush_mean = rbind(bush_mean, data.frame(yr,mn, mean3))
      }
    }
    
    bush_mean <- bush_mean[with(bush_mean, order(mn,yr)), ]
    
    
    ###################################################################################################
    
    #mean of weights by month for each year for Chestnut-backed Chickadee 
    cbch_mean = data.frame()
    
    for ( yr in unique(cbch$year) )
    {
      for ( mn in unique(cbch$month) )
      {
        tmp4 = subset(cbch, year==yr & month==mn)
        mean4 = mean(tmp4$cleanWeight, na.rm=TRUE)
        cbch_mean = rbind(cbch_mean, data.frame(yr,mn, mean4))
      }
    }
    
    cbch_mean <- cbch_mean[with(cbch_mean, order(mn,yr)), ]
    
    ####################################################################################################
    
    k <- subset(sosp_mean, yr=="2016"|yr=="2015"|yr=="2014"|yr=="2013"|yr=="2012")
    k$yr<-factor(k$yr)
    ggplot(data = k, aes(x = mn, y = mean1, color=yr)) + 
      geom_line()+ geom_point()+
      xlab("Month")+ylab("Weight")+labs(title="Weight changes per month for Song Sparrow",colour="Year")+
      theme(plot.title = element_text(hjust = 0.5))+scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12))
    
    
    l <- subset(coye_mean, yr=="2016"|yr=="2015"|yr=="2014"|yr=="2013"|yr=="2012")
    l$yr<-factor(l$yr)
    ggplot(data = l, aes(x = mn, y = mean, color=yr)) + 
      geom_line()+ geom_point()+
      xlab("Month")+ylab("Weight")+labs(title="Weight changes per month for Common Yellowthroat",colour="Year")+
      theme(plot.title = element_text(hjust = 0.5))+scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12))
    
    m <- subset(bewr_mean, yr=="2016"|yr=="2015"|yr=="2014"|yr=="2013"|yr=="2012")
    m$yr<-factor(m$yr)
    ggplot(data = m, aes(x = mn, y = mean2, color=yr)) + 
      geom_line()+ geom_point()+
      xlab("Month")+ylab("Weight")+labs(title="Weight changes per month for Bewick's Wren",colour="Year")+
      theme(plot.title = element_text(hjust = 0.5))+scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12))
    
    n <- subset(bush_mean, yr=="2016"|yr=="2015"|yr=="2014"|yr=="2013"|yr=="2012")
    n$yr<-factor(n$yr)
    ggplot(data = n, aes(x = mn, y = mean3, color=yr)) + 
      geom_line()+ geom_point()+
      xlab("Month")+ylab("Weight")+labs(title="Weight changes per month for Bushtit",colour="Year")+
      theme(plot.title = element_text(hjust = 0.5))+scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12))
    
    o <- subset(cbch_mean, yr=="2016"|yr=="2015"|yr=="2014"|yr=="2013"|yr=="2012")
    o$yr<-factor(o$yr)
    ggplot(data = o, aes(x = mn, y = mean4, color=yr)) + 
      geom_line()+ geom_point()+
      xlab("Month")+ylab("Weight")+labs(title="Weight changes per month for Chestnut-backed Chickadee",colour="Year")+
      theme(plot.title = element_text(hjust = 0.5))+scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12))
    
    ```
    
#### Change in the average Wing to Weight Ratio of each species over the span of 20 years
* The average wing to weight ration of all the species is constant over the past 20 years.

    ```{r, echo=FALSE}
    library(dplyr)
    library(ggplot2)
    #summary statistics based on years for Common Yellowthroat for wingtoweight ratio
    p=coye%>%
      group_by(year)%>% 
      summarise(Mean=mean(wingToWeightRatio, na.rm=TRUE),Max=max(wingToWeightRatio, na.rm=TRUE), Min=min(wingToWeightRatio, na.rm=TRUE), Median=median(wingToWeightRatio, na.rm=TRUE), Std=sd(wingToWeightRatio, na.rm=TRUE))
    
    #summary statistics based on years for Song Sparrow for wingtoweight ratio
    q=sosp%>%
      group_by(year)%>% 
      summarise(Mean=mean(wingToWeightRatio, na.rm=TRUE),Max=max(wingToWeightRatio, na.rm=TRUE), Min=min(wingToWeightRatio, na.rm=TRUE), Median=median(wingToWeightRatio, na.rm=TRUE), Std=sd(wingToWeightRatio, na.rm=TRUE))
    
    #summary statistics based on years for Bewick's Wren for wingtoweight ratio
    r=bewr%>%
      group_by(year)%>% 
      summarise(Mean=mean(wingToWeightRatio, na.rm=TRUE),Max=max(wingToWeightRatio, na.rm=TRUE), Min=min(wingToWeightRatio, na.rm=TRUE), Median=median(wingToWeightRatio, na.rm=TRUE), Std=sd(wingToWeightRatio, na.rm=TRUE))
    
    #summary statistics based on years for Bushtit for wingtoweight ratio
    s=bush%>%
      group_by(year)%>% 
      summarise(Mean=mean(wingToWeightRatio, na.rm=TRUE),Max=max(wingToWeightRatio, na.rm=TRUE), Min=min(wingToWeightRatio, na.rm=TRUE), Median=median(wingToWeightRatio, na.rm=TRUE), Std=sd(wingToWeightRatio, na.rm=TRUE))
    
    #summary statistics based on years for Chestnut-backed Chickadee for wingtoweight ratio
    t=cbch%>%
      group_by(year)%>% 
      summarise(Mean=mean(wingToWeightRatio, na.rm=TRUE),Max=max(wingToWeightRatio, na.rm=TRUE), Min=min(wingToWeightRatio, na.rm=TRUE), Median=median(wingToWeightRatio, na.rm=TRUE), Std=sd(wingToWeightRatio, na.rm=TRUE))
    
    ############################################################################################################################################
    
    ggplot() + 
      geom_line(data=p, aes(x=year, y=Mean, color="Song sparrow"),size=1) +
      geom_line(data=q, aes(x=year, y=Mean, color="Common Yellow Throat"),size=1)+
      geom_line(data=r, aes(x=year, y=Mean, color="Bewick's Wren"),size=1) +
      geom_line(data=s, aes(x=year, y=Mean, color="Bushtit"),size=1)+
      geom_line(data=t, aes(x=year, y=Mean, color="Chestnut-backed Chickadee"),size=1)+xlab("Year")+ylab("Mean of Wing to Weight Ratios")+labs(title="Wing to Weight Ratio changes by Year",colour="Species Name")+
      theme(plot.title = element_text(hjust = 0.5))
    
    ```
    
#### Change in the average Wing to Weight Ratio of each species by month over the span of 5 years
* The pattern of changes in the wing to weight ratio per month is different for each species, but is fairly constant with a few outliers between 2012 to 2016 

    ```{r, echo=FALSE}
    
    #mean of wing to weight by month for each year for Common Yellowthroat
    coye_wtow_mean = data.frame()
    
    for ( yr in unique(coye$year) )
    {
      for ( mn in unique(coye$month) )
      {
        tmp5 = subset(coye, year==yr & month==mn)
        mean_c_wtow = mean(tmp5$wingToWeightRatio)
        coye_wtow_mean = rbind(coye_wtow_mean, data.frame(yr,mn, mean_c_wtow))
      }
    }
    
    coye_wtow_mean <- coye_wtow_mean[with(coye_wtow_mean, order(mn,yr)), ]
    
    ####################################################################################################
    
    #mean of wing to weight by month for each year for Song Sparrow
    sosp_wtow_mean = data.frame()
    
    for ( yr in unique(sosp$year) )
    {
      for ( mn in unique(sosp$month) )
      {
        tmp6 = subset(sosp, year==yr & month==mn)
        mean_s_wtow = mean(tmp6$wingToWeightRatio)
        sosp_wtow_mean = rbind(sosp_wtow_mean, data.frame(yr,mn, mean_s_wtow))
      }
    }
    
    sosp_wtow_mean <- sosp_wtow_mean[with(sosp_wtow_mean, order(mn,yr)), ]
    
    ####################################################################################################
    
    #mean of wing to weight by month for each year for Bewick's Wren
    bewr_wtow_mean = data.frame()
    
    for ( yr in unique(bewr$year) )
    {
      for ( mn in unique(bewr$month) )
      {
        tmp7 = subset(bewr, year==yr & month==mn)
        mean_be_wtow = mean(tmp7$wingToWeightRatio)
        bewr_wtow_mean = rbind(bewr_wtow_mean, data.frame(yr,mn, mean_be_wtow))
      }
    }
    
    bewr_wtow_mean <- bewr_wtow_mean[with(bewr_wtow_mean, order(mn,yr)), ]
    
    ####################################################################################################
    
    #mean of wing to weight by month for each year for Bushtit
    bush_wtow_mean = data.frame()
    
    for ( yr in unique(bush$year) )
    {
      for ( mn in unique(bush$month) )
      {
        tmp8 = subset(bush, year==yr & month==mn)
        mean_bu_wtow = mean(tmp8$wingToWeightRatio)
        bush_wtow_mean = rbind(bush_wtow_mean, data.frame(yr,mn, mean_bu_wtow))
      }
    }
    
    bush_wtow_mean <- bush_wtow_mean[with(bush_wtow_mean, order(mn,yr)), ]
    
    ####################################################################################################
    
    #mean of wing to weight by month for each year for Chestnut-backed Chickadee
    cbch_wtow_mean = data.frame()
    
    for ( yr in unique(cbch$year) )
    {
      for ( mn in unique(cbch$month) )
      {
        tmp9 = subset(cbch, year==yr & month==mn)
        mean_cb_wtow = mean(tmp9$wingToWeightRatio)
        cbch_wtow_mean = rbind(cbch_wtow_mean, data.frame(yr,mn, mean_cb_wtow))
      }
    }
    
    cbch_wtow_mean <- cbch_wtow_mean[with(cbch_wtow_mean, order(mn,yr)), ]
    
    ########################################################################################################
    
    u <- subset(sosp_wtow_mean, yr=="2016"|yr=="2015"|yr=="2014"|yr=="2013"|yr=="2012")
    u$yr<-factor(u$yr)
    ggplot(data = u, aes(x = mn, y = mean_s_wtow, color=yr)) + 
      geom_line()+ geom_point()+
      xlab("Month")+ylab("Wing to Weight Ratio")+labs(title="Change in Wing to weight Ratio per month for Song Sparrow",colour="Year")+
      theme(plot.title = element_text(hjust = 0.5))+scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12))
    
    v <- subset(coye_wtow_mean, yr=="2016"|yr=="2015"|yr=="2014"|yr=="2013"|yr=="2012")
    v$yr<-factor(v$yr)
    ggplot(data = v, aes(x = mn, y = mean_c_wtow, color=yr)) + 
      geom_line()+ geom_point()+
      xlab("Month")+ylab("Wing to Weight Ratio")+labs(title="Change in Wing to weight Ratio per month for Common Yellowthroat",colour="Year")+
      theme(plot.title = element_text(hjust = 0.5))+scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12))
    
    w <- subset(bewr_wtow_mean, yr=="2016"|yr=="2015"|yr=="2014"|yr=="2013"|yr=="2012")
    w$yr<-factor(w$yr)
    ggplot(data = w, aes(x = mn, y = mean_be_wtow, color=yr)) + 
      geom_line()+ geom_point()+
      xlab("Month")+ylab("Wing to Weight Ratio")+labs(title="Change in Wing to weight Ratio per month for Bewick's Wren",colour="Year")+
      theme(plot.title = element_text(hjust = 0.5))+scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12))
    
    x <- subset(bush_wtow_mean, yr=="2016"|yr=="2015"|yr=="2014"|yr=="2013"|yr=="2012")
    x$yr<-factor(x$yr)
    ggplot(data = x, aes(x = mn, y = mean_bu_wtow, color=yr)) + 
      geom_line()+ geom_point()+
      xlab("Month")+ylab("Wing to Weight Ratio")+labs(title="Change in Wing to weight Ratio per month for Bushtit",colour="Year")+
      theme(plot.title = element_text(hjust = 0.5))+scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12))
    
    y <- subset(cbch_wtow_mean, yr=="2016"|yr=="2015"|yr=="2014"|yr=="2013"|yr=="2012")
    y$yr<-factor(y$yr)
    ggplot(data = y, aes(x = mn, y = mean_cb_wtow, color=yr)) + 
      geom_line()+ geom_point()+
      xlab("Month")+ylab("Wing to Weight Ratio")+labs(title="Change in Wing to weight Ratio per month for Chestnut-backed Chickadee",colour="Year")+
      theme(plot.title = element_text(hjust = 0.5))+scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12))

    ```
    
## Contributorship Statement




## Proofreader